What paradigms can webcam eye-tracking be used for? Attempted replications of five ``classic'' cognitive science experiments

### 2025-03-30 manuscript.Rmd
Switched to visual tab in R studio

While promising, there are some salient limitations. First, many of the studies report effects that are smaller or later than what had been previously observed in the lab [@slim2024webcams; @van2024validation; @slim2022moving; @degen2021seeing; @van2023validation; @bogdan2024investigating; @kandel2024assessing]. Some of this is likely due to programming errors rather than inherent limitations in the technology: Importantly, accurate timing on a web browser is not trivial [@passell2021cognitive; @de2016psychophysics], and subtle programming choices can significantly affect the accuracy of `WebGazer.js` timing [@yang2021webcam].^[See also discussion at https://github.com/jspsych/jsPsych/discussions/1892] Since most of the prior work did not address these timing issues, it is not clear how many of the reported lab/web differences would resolve.

@@

All data and analysis code underlying this paper can be found in the Github repository: https://github.com/jodeleeuw/219-2021-eyetracking-analysis. The current paper is fully reproducible by rendering `manuscript.Rmd`.

@@

We used a fork of the `webgazer.js` library for webcam eye-tracking
[@papoutsaki2016webgazer], implemented in `jsPsych`, a Javascript
library for running behavioral experiments in a web browser
[@deleeuwJsPsychJavaScriptLibrary2015].[^1] Our fork included changes to
`webgazer.js` in order to improve data quality for experiments in which
the precise timing of stimulus onsets is relevant. Specifically, we
implemented a polling mode so that gaze predictions could be requested
at a regular interval, which improved the sampling rate considerably in
informal testing. This modification is similar to what Yang and Krajbich
[-@yang2021webcam] reported improved the sampling rate in their study of
webgazer. We also adjusted the mechanism for recording time stamps of
each gaze prediction, so that the time stamp reported by webgazer is
based on when the video frame is received and not when the computation
of the gaze point is finished.

@@

The first study was a replication attempt of
@altmannIncrementalInterpretationVerbs1999. Altmann and Kamide used the
visual world eye-tracking paradigm [@tanenhaus1995] to show that
meanings of verbs rapidly constrain the set of potential subsequent
referents in sentence processing. For example, when looking at the
display in Figure\ \@ref(fig:E1-example-trial) and listening to a sentence like
"The boy will eat the...," participants are more likely to look at the cake than 
when they hear "The boy will move the...," in which case they tend to look at the
train, presumably because cakes are edible and trains are not. Semantic
information available at the verb is used to anticipate upcoming
linguistic input.

@@

### Participants
Sixty participants (2.5 times the original sample of 24) were paid \$2.60 for their participation. For unknown reasons,
2 of the subjects' results were not recorded, so in the analysis, we
worked with data collected from 58 participants.

@@

The visual stimuli were created through Canva and depicted an agent
accompanied by four to five objects in the scene (see Figure\ \@ref(fig:E1-example-trial)). On
critical trials, participants heard one of two sentences associated with
the scene. In the restrictive condition, the sentence (e.g., "The boy
will eat the cake") contained a verb (e.g., "eat") which restricts the
set of possible subsequent referents (e.g., to edible things). Only the
target object (e.g., the cake) was semantically consistent with the
verb's meaning. In the non-restrictive condition, the sentence (e.g.,
"The boy will move the cake") contained a verb (e.g., "move") which does
not restrict the set of possible subsequent referents. The target object
(e.g., the cake) as well as the distractor objects (e.g., the train, the
ball, etc.) were semantically consistent with the verb's meaning. Both
sentences were compatible with the scene, such that the correct keyboard
response for the critical trials was "yes." Filler trials consisted of
scenes that also contained an agent surrounded by objects as in the critical trials, but corresponding sentences named an object that was not present in the scene. The correct keyboard response for the filler
trials was "no."

@@

```{=html}
<!-- Qs: 
-keyboard keys? what was the exact phrasing of the instruction? 
-Any delay between when visual scene appears and the audio starts? 
-Does trial end as soon as participant responds?-->
```

## Results
Looks to the objects in the scene were time-locked to the onset of the
verb, the offset of the verb, onset of the post-verbal determiner, and
onset of the target noun. ROIs were defined by creating boxes around each object in the scene. The size of each box was determined by taking the height and width of the given object and adding 20 pixels of padding. Each scene contained an agent region, a target region, and three or four distractor regions.

@@

## Discussion
Across three different tests of the hypothesis that listeners will use verb semantics to anticipate the upcoming referent, we found results that were numerically consistent with the hypothesis but not statistically significant. This is unlikely to be because the original effect is unreliable. Indeed, one of us had successfully replicated it previously using traditional methods [@james2023language].

@@

# Experiment 1b
Experiment 1b tested whether the failure to replicate significant condition effects in Experiment 1a was due to features of conducting the study remotely (i.e. varied experimental settings and apparatuses, lower compliance) rather than webcam-based eye-tracking or `webgazer` _per se_. Thus, Experiment 1b took place in a lab setting with undergraduate participants but otherwise used the same Method as Experiment 1a.

@@

### Participants
Forty-nine participants completed the study in a lab setting. They were recruited via the Boston College subject pool. Participants needed to be 18 years of age or older and native speakers of English.

### Materials and Design
Materials were identical to those in Experiment 1a.

### Procedure
After being greeted by the experimenter and completing the informed consent form, participants followed on-screen prompts to complete the study, including calibration, as described in the Experiment 1a Procedure.

### Apparatus
Subjects were tested on a MacIntosh laptop using the built-in Webcam. The study was run in Chrome using the same code as Exp. 1a.

## Results
```{r E1-analysis-JoshH-lab, child = 'group-a/E1-analysis-comparison-to-Hartshorne.Rmd'}

@@

## Discussion
Overall, the results of Experiment 1 paint a sobering picture of web-based eye-tracking. In Experiment 1a, results from the remote sample were in the expected direction but effects were smaller and delayed relative to previous work and failed to reach statistical significance. This was not due to vagaries of remote samples or working with a more diverse subject population: results from the in-lab study were, if anything, worse. Nor does our failure to replicate seem to be driven by subjects with poor calibration; effect sizes were not correlated with calibration success, and restricting analyses to subjects with good calibration scores only helped in one of the experiments and in only one of the three critical analyses.

@@

The experiment consisted of two blocks each composed of an encoding
phase and a recall phase. The two blocks differed by whether the recall phase was in the free-viewing or fixed-viewing condition, as described in the Procedure. Participants were randomly assigned to see the
fixed-viewing or free-viewing block first. 
During each encoding phase, participants saw a
grid indicating the four quadrants of the screen. Each quadrant
contained six cartoon images of items belonging to the same category. The four categories were humanoids, household objects, animals, and
methods of transportation (see Figure\ \@ref(fig:E2-example-trial)). Different images were used in each block; there were 48 unique images total across the experiment.

@@

Stimulus sets were created to present participants with 24 presentation trials and 24 test trials on Day 1, and 48 recognition trials on Day 2. Each presentation trial screen presented two identical images (e.g. red apple and red apple). The test trial screens presented each of the previously-seen ("old") images paired with a new image (e.g. red apple and bicycle key). The recognition trial screens presented a single image that was either from the original set of 24 (seen twice across the first two phases) or was a corresponding foil for the original set (e.g. red apple or green apple). Thus, each participant was exposed to 72 unique images over the course of the experiment (the 24 foils for the "new" images in the test phase are never seen). Two full stimulus lists were created to counterbalance images across participants; the images that composed the "old" set for one list made up the "new" set for the list. The experimental design is visually depicted in Figure\ \@ref(fig:E3-design-schematic).

@@

During a trial, four pictures were displayed (target animal, target
instrument, distractor animal, distractor instrument), one in each
corner of the screen, and participants heard an audio prompt that
contained instructions about the action they needed to act out (e.g.,
"Rub the butterfly with the crayon"; see Figure\ \@ref(fig:E4-example-trial))[^2]. Using their
cursor, participants could act out the instructions by clicking on
objects and moving them or motioning over the objects[^3]. After the
action was completed, the participants were instructed to press the
space bar which led to a screen that said "Click Here" in the middle in
order to remove bias in the eye and mouse movements from the previous
trial. The experiment only allowed the participants to move on to the
next trial once the audio was completely done playing and the mouse had
been moved over at least one object.

[^1]: The WebGazer.js patch can be found at https://github.com/brownhci/WebGazer/commit/35301e56c192ef265bb270873a6526205f183711.

[^2]: In the original study, the pictures appeared one by one on the
    screen and their names were played as they appeared. We removed this
    introductory portion of the trial to save time

[^3]:  As opposed to the original study we recorded mouse movement
    instead of clicking behavior since not all of the audio prompts
    required clicking. For example, the sentence "locate the camel with
    the straw" may not involve any clicking but rather only mousing over
    the camel.

@@

All stimuli, experiment scripts, data, and analysis scripts are
available on the Open Science Framework at [https://osf.io/eubsc/](https://osf.io/eubsc/). 
The study pre-registration is available at
[https://osf.io/tv57s](https://osf.io/tv57s).

@@

During each trial of the main task, two faces were displayed on the two
halves of the screen, one on the left and one on the right (as in Figure\ \@ref(fig:E5-example-trial)). In the attractiveness task,
participants were asked to chose the more attractice face in the pair
and in the shape judgment task participants were asked to pick the face
that appeared rounder. They pressed the "a" key on their keyboard to
select the face on the left and the "d" key to select the face on the
right. A fixation cross appeared in the center of the screen between
each set of faces. Participants were asked to look at this fixation
cross in order to reset their gaze in between trials. The order of
the 19 face pairs was random for each participant.

@@

## Results
In the original study, a video-based eye tracker was used. The eye
movements of participants were recorded with a digital camera
downsampled to 33.3 Hz, with eye position determined
automatically with MediaAnalyzer software. In our study, subjects
supplied their own cameras, so hardware sampling rate varied. However,
data was collected at 20 Hz. <!--`r colorize("[TODO - CONFIRM]", 'red')` AJ: I don't know how to confirm this! -->

@@

 -->
```

# General Discussion

@@

Second, researchers need to take care with timing accuracy. Running studies on modern computers requires synchronizing multiple clocks, something that does not happen automatically. Failure to synchronize can result in substantial inaccuracies [@slim2022moving]. As indicated by the current work, the current jsPsych implementation of WebGazer.js has timing synchronization sufficient for many studies. We urge researchers to avoid using the current release of WebGazer.js as is, and instead either use jsPsych, our patch to WebGazer.js (https://github.com/brownhci/WebGazer/commit/35301e56c192ef265bb270873a6526205f183711), or other software that has been specifically documented as addressing common issues in timing for web-based eye-tracking.

@@

# Contributions
* Joshua R. de Leeuw
    + Conceptualization
    + Data curation
    + Formal analysis
    + Investigation
    + Methodology
    + Project administration
    + Software
    + Supervision
    + Validation
    + Visualization
    + Writing - original draft
    + Writing - review & editing

* Ariel N. James, Rachel Ryskin, and Joshua K. Hartshorne
    + Conceptualization
    + Formal analysis
    + Visualization
    + Writing - original draft
    + Writing - review & editing

* Undergraduate student authors, listed alphabetically: Haylee Backs, Nandeeta Bala, Laila Barcenas-Meade, Samata Bhattarai, Tessa Charles, Gerasimos Copoulos, Claire Coss, Alexander Eisert, Elena Furuhashi, Keara Ginell, Anna Guttman-McCabe, Emma (Chaz) Harrison, Laura Hoban, William A. Hwang, Claire Iannetta, Kristen M. Koenig, Chauncey Lo, Victoria Palone, Gina Pepitone, Margaret Ritzau, Yi Hua Sung, and Lauren Thompson
    + Investigation
    + Methodology
    + Software

@@

# Funding information
Funding for data collection was provided by the Vassar College Department of Cognitive Science (JdL). Partial funding for JKH's contributions was provided by NSF2436092.

@@

# Competing interests
The authors have no competing interests to declare.

@@

# Data accessibility statement
All data and analysis code underlying this paper can be found in the Github repository: https://github.com/jodeleeuw/219-2021-eyetracking-analysis. Study materials and pre-registrations for each experiment are available on Open Science Framework:

* Experiments 1a and 1b: [https://osf.io/g6dum/]
* Experiment 2: [https://osf.io/xezfu/]
* Experiment 3: [https://osf.io/k63b9/]
* Experiment 4: [https://osf.io/x3c49/]
* Experiment 5: [https://osf.io/eubsc/]

>>>>

While promising, there are some salient limitations. First, many of the studies report effects that are smaller or later than what had been previously observed in the lab [@slim2024webcams; @van2024validation; @slim2022moving; @degen2021seeing; @van2023validation; @bogdan2024investigating; @kandel2024assessing]. Some of this is likely due to programming errors rather than inherent limitations in the technology: Importantly, accurate timing on a web browser is not trivial [@passell2021cognitive; @de2016psychophysics], and subtle programming choices can significantly affect the accuracy of `WebGazer.js` timing [@yang2021webcam].[^1] Since most of the prior work did not address these timing issues, it is not clear how many of the reported lab/web differences would resolve.

[^1]: See also discussion at <https://github.com/jspsych/jsPsych/discussions/1892>

@@

All data and analysis code underlying this paper can be found in the Github repository: <https://github.com/jodeleeuw/219-2021-eyetracking-analysis>. The current paper is fully reproducible by rendering `manuscript.Rmd`.

@@

We used a fork of the `webgazer.js` library for webcam eye-tracking [@papoutsaki2016webgazer], implemented in `jsPsych`, a Javascript library for running behavioral experiments in a web browser [@deleeuwJsPsychJavaScriptLibrary2015].[^2] Our fork included changes to `webgazer.js` in order to improve data quality for experiments in which the precise timing of stimulus onsets is relevant. Specifically, we implemented a polling mode so that gaze predictions could be requested at a regular interval, which improved the sampling rate considerably in informal testing. This modification is similar to what Yang and Krajbich [-@yang2021webcam] reported improved the sampling rate in their study of webgazer. We also adjusted the mechanism for recording time stamps of each gaze prediction, so that the time stamp reported by webgazer is based on when the video frame is received and not when the computation of the gaze point is finished.

[^2]: The WebGazer.js patch can be found at <https://github.com/brownhci/WebGazer/commit/35301e56c192ef265bb270873a6526205f183711>.

@@

The first study was a replication attempt of @altmannIncrementalInterpretationVerbs1999. Altmann and Kamide used the visual world eye-tracking paradigm [@tanenhaus1995] to show that meanings of verbs rapidly constrain the set of potential subsequent referents in sentence processing. For example, when looking at the display in Figure \@ref(fig:E1-example-trial) and listening to a sentence like "The boy will eat the...," participants are more likely to look at the cake than when they hear "The boy will move the...," in which case they tend to look at the train, presumably because cakes are edible and trains are not. Semantic information available at the verb is used to anticipate upcoming linguistic input.

@@

Sixty participants (2.5 times the original sample of 24) were paid \$2.60 for their participation. For unknown reasons, 2 of the subjects' results were not recorded, so in the analysis, we worked with data collected from 58 participants.

@@

The visual stimuli were created through Canva and depicted an agent accompanied by four to five objects in the scene (see Figure \@ref(fig:E1-example-trial)). On critical trials, participants heard one of two sentences associated with the scene. In the restrictive condition, the sentence (e.g., "The boy will eat the cake") contained a verb (e.g., "eat") which restricts the set of possible subsequent referents (e.g., to edible things). Only the target object (e.g., the cake) was semantically consistent with the verb's meaning. In the non-restrictive condition, the sentence (e.g., "The boy will move the cake") contained a verb (e.g., "move") which does not restrict the set of possible subsequent referents. The target object (e.g., the cake) as well as the distractor objects (e.g., the train, the ball, etc.) were semantically consistent with the verb's meaning. Both sentences were compatible with the scene, such that the correct keyboard response for the critical trials was "yes." Filler trials consisted of scenes that also contained an agent surrounded by objects as in the critical trials, but corresponding sentences named an object that was not present in the scene. The correct keyboard response for the filler trials was "no."

@@

```{=html}
<!-- Qs: 
-keyboard keys? what was the exact phrasing of the instruction? 
-Any delay between when visual scene appears and the audio starts? 
-Does trial end as soon as participant responds?-->
```
## Results

Looks to the objects in the scene were time-locked to the onset of the verb, the offset of the verb, onset of the post-verbal determiner, and onset of the target noun. ROIs were defined by creating boxes around each object in the scene. The size of each box was determined by taking the height and width of the given object and adding 20 pixels of padding. Each scene contained an agent region, a target region, and three or four distractor regions.

@@

Across three different tests of the hypothesis that listeners will use verb semantics to anticipate the upcoming referent, we found results that were numerically consistent with the hypothesis but not statistically significant. This is unlikely to be because the original effect is unreliable. Indeed, one of us had successfully replicated it previously using traditional methods [@james2023language].

@@

# Experiment 1b

Experiment 1b tested whether the failure to replicate significant condition effects in Experiment 1a was due to features of conducting the study remotely (i.e. varied experimental settings and apparatuses, lower compliance) rather than webcam-based eye-tracking or `webgazer` *per se*. Thus, Experiment 1b took place in a lab setting with undergraduate participants but otherwise used the same Method as Experiment 1a.

@@

Forty-nine participants completed the study in a lab setting. They were recruited via the Boston College subject pool. Participants needed to be 18 years of age or older and native speakers of English.

Materials were identical to those in Experiment 1a.

After being greeted by the experimenter and completing the informed consent form, participants followed on-screen prompts to complete the study, including calibration, as described in the Experiment 1a Procedure.

### Apparatus

Subjects were tested on a MacIntosh laptop using the built-in Webcam. The study was run in Chrome using the same code as Exp. 1a.

```{r E1-analysis-JoshH-lab, child = 'group-a/E1-analysis-comparison-to-Hartshorne.Rmd'}

@@

Overall, the results of Experiment 1 paint a sobering picture of web-based eye-tracking. In Experiment 1a, results from the remote sample were in the expected direction but effects were smaller and delayed relative to previous work and failed to reach statistical significance. This was not due to vagaries of remote samples or working with a more diverse subject population: results from the in-lab study were, if anything, worse. Nor does our failure to replicate seem to be driven by subjects with poor calibration; effect sizes were not correlated with calibration success, and restricting analyses to subjects with good calibration scores only helped in one of the experiments and in only one of the three critical analyses.

@@

The experiment consisted of two blocks each composed of an encoding phase and a recall phase. The two blocks differed by whether the recall phase was in the free-viewing or fixed-viewing condition, as described in the Procedure. Participants were randomly assigned to see the fixed-viewing or free-viewing block first. During each encoding phase, participants saw a grid indicating the four quadrants of the screen. Each quadrant contained six cartoon images of items belonging to the same category. The four categories were humanoids, household objects, animals, and methods of transportation (see Figure \@ref(fig:E2-example-trial)). Different images were used in each block; there were 48 unique images total across the experiment.

@@

Stimulus sets were created to present participants with 24 presentation trials and 24 test trials on Day 1, and 48 recognition trials on Day 2. Each presentation trial screen presented two identical images (e.g. red apple and red apple). The test trial screens presented each of the previously-seen ("old") images paired with a new image (e.g. red apple and bicycle key). The recognition trial screens presented a single image that was either from the original set of 24 (seen twice across the first two phases) or was a corresponding foil for the original set (e.g. red apple or green apple). Thus, each participant was exposed to 72 unique images over the course of the experiment (the 24 foils for the "new" images in the test phase are never seen). Two full stimulus lists were created to counterbalance images across participants; the images that composed the "old" set for one list made up the "new" set for the list. The experimental design is visually depicted in Figure \@ref(fig:E3-design-schematic).

@@

During a trial, four pictures were displayed (target animal, target instrument, distractor animal, distractor instrument), one in each corner of the screen, and participants heard an audio prompt that contained instructions about the action they needed to act out (e.g., "Rub the butterfly with the crayon"; see Figure \@ref(fig:E4-example-trial))[^3]. Using their cursor, participants could act out the instructions by clicking on objects and moving them or motioning over the objects[^4]. After the action was completed, the participants were instructed to press the space bar which led to a screen that said "Click Here" in the middle in order to remove bias in the eye and mouse movements from the previous trial. The experiment only allowed the participants to move on to the next trial once the audio was completely done playing and the mouse had been moved over at least one object.

[^3]: In the original study, the pictures appeared one by one on the screen and their names were played as they appeared. We removed this introductory portion of the trial to save time

[^4]: As opposed to the original study we recorded mouse movement instead of clicking behavior since not all of the audio prompts required clicking. For example, the sentence "locate the camel with the straw" may not involve any clicking but rather only mousing over the camel.

@@

All stimuli, experiment scripts, data, and analysis scripts are available on the Open Science Framework at <https://osf.io/eubsc/>. The study pre-registration is available at <https://osf.io/tv57s>.

@@

During each trial of the main task, two faces were displayed on the two halves of the screen, one on the left and one on the right (as in Figure \@ref(fig:E5-example-trial)). In the attractiveness task, participants were asked to chose the more attractice face in the pair and in the shape judgment task participants were asked to pick the face that appeared rounder. They pressed the "a" key on their keyboard to select the face on the left and the "d" key to select the face on the right. A fixation cross appeared in the center of the screen between each set of faces. Participants were asked to look at this fixation cross in order to reset their gaze in between trials. The order of the 19 face pairs was random for each participant.

@@

In the original study, a video-based eye tracker was used. The eye movements of participants were recorded with a digital camera downsampled to 33.3 Hz, with eye position determined automatically with MediaAnalyzer software. In our study, subjects supplied their own cameras, so hardware sampling rate varied. However, data was collected at 20 Hz. <!--`r colorize("[TODO - CONFIRM]", 'red')` AJ: I don't know how to confirm this! -->

@@

 -->
```
# General Discussion

@@

Second, researchers need to take care with timing accuracy. Running studies on modern computers requires synchronizing multiple clocks, something that does not happen automatically. Failure to synchronize can result in substantial inaccuracies [@slim2022moving]. As indicated by the current work, the current jsPsych implementation of WebGazer.js has timing synchronization sufficient for many studies. We urge researchers to avoid using the current release of WebGazer.js as is, and instead either use jsPsych, our patch to WebGazer.js (<https://github.com/brownhci/WebGazer/commit/35301e56c192ef265bb270873a6526205f183711>), or other software that has been specifically documented as addressing common issues in timing for web-based eye-tracking.

@@

# Contributions

-   Joshua R. de Leeuw
    -   Conceptualization
    -   Data curation
    -   Formal analysis
    -   Investigation
    -   Methodology
    -   Project administration
    -   Software
    -   Supervision
    -   Validation
    -   Visualization
    -   Writing - original draft
    -   Writing - review & editing
-   Ariel N. James, Rachel Ryskin, and Joshua K. Hartshorne
    -   Conceptualization
    -   Formal analysis
    -   Visualization
    -   Writing - original draft
    -   Writing - review & editing
-   Undergraduate student authors, listed alphabetically: Haylee Backs, Nandeeta Bala, Laila Barcenas-Meade, Samata Bhattarai, Tessa Charles, Gerasimos Copoulos, Claire Coss, Alexander Eisert, Elena Furuhashi, Keara Ginell, Anna Guttman-McCabe, Emma (Chaz) Harrison, Laura Hoban, William A. Hwang, Claire Iannetta, Kristen M. Koenig, Chauncey Lo, Victoria Palone, Gina Pepitone, Margaret Ritzau, Yi Hua Sung, and Lauren Thompson
    -   Investigation
    -   Methodology
    -   Software

@@

# Funding information

Funding for data collection was provided by the Vassar College Department of Cognitive Science (JdL). Partial funding for JKH's contributions was provided by NSF2436092.

@@

# Competing interests

The authors have no competing interests to declare.

@@

# Data accessibility statement

All data and analysis code underlying this paper can be found in the Github repository: <https://github.com/jodeleeuw/219-2021-eyetracking-analysis>. Study materials and pre-registrations for each experiment are available on Open Science Framework:

-   Experiments 1a and 1b: [<https://osf.io/g6dum/>]
-   Experiment 2: [<https://osf.io/xezfu/>]
-   Experiment 3: [<https://osf.io/k63b9/>]
-   Experiment 4: [<https://osf.io/x3c49/>]
-   Experiment 5: [<https://osf.io/eubsc/>]
